\chapter{User Manual\label{chap:user-manual}}
\lstset{style=mystyle}
\section{Prerequisites}
\begin{itemize}
    \item Python 3.11.11
\end{itemize}

\section{Installation}
\subsection{Step 1: Clone or Unzip the Repository}
Clone the repository or unzip the \texttt{counterfactual\_explainers-main.zip} file.
\begin{lstlisting}[style=shell]
git clone https://github.com/Hariss-Gills/counterfactual_explainers.git
\end{lstlisting}

\subsection{Step 2: Create Virtual Environments}
Create virtual environments for each counterfactual method.
\begin{lstlisting}[style=shell]
python -m venv ~/.python-venvs/counterfactual-explainers-dice
python -m venv ~/.python-venvs/counterfactual-explainers-aide
\end{lstlisting}

\subsection{Step 3: Install Required Packages}
Install the necessary packages for each method.
\begin{lstlisting}[style=shell]
cd counterfactual_explainers

source ~/.python-venvs/counterfactual-explainers-dice/bin/activate
pip install .
pip install -r requirements-dice.txt
deactivate

source ~/.python-venvs/counterfactual-explainers-aide/bin/activate
pip install .
pip install -r requirements-aide.txt
deactivate
\end{lstlisting}

\section{Usage}
Run the top level modules using the respective virtual environments.
\begin{lstlisting}[style=shell]
source ~/.python-venvs/counterfactual-explainers-dice/bin/activate
python counterfactual_explainers/dataset_stats.py
python counterfactual_explainers/train_models.py
python counterfactual_explainers/gen_cfs_dice.py
deactivate

source ~/.python-venvs/counterfactual-explainers-aide/bin/activate
python counterfactual_explainers/gen_cfs_aide.py
python counterfactual_explainers/calculate_metrics.py
python counterfactual_explainers/plot_and_stats.py
deactivate
\end{lstlisting}

\section{Notes}
\begin{itemize}
    \item It is necessary to downgrade from Python 3.13.1 to 3.12.8 due to TensorFlow compatibility.
    \item TensorFlow must be downgraded to \texttt{tensorflow==2.13.0} since \texttt{dice\_ml==0.11} uses that version. Downgrading scikit-learn might also resolve the deserialization issue between using the pipeline versus a raw model.
    \item Since \href{https://github.com/scikit-learn/scikit-learn/blob/d666202a9349893c1bd106cc9ee0ff0a807c7cf3/sklearn/base.py#L540}{\texttt{ClassifierMixin}} does not have a superclass, the \texttt{Tags()} class is used to instantiate some tags instead of downgrading scikit-learn.
\end{itemize}